Landform boundary effects on Holocene forager landscape use in arid South Australia

### 2017-01-26 analysis/paper/paper.Rmd
got the text tables in

Philip Hughes    
Department of Archaeology, Flinders University, GPO Box 2100, Adelaide, SA, 5001

Marjorie Sullivan    
The Australian National University, Canberra ACT 0200, Australia and 
Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia

@@

A key step in understanding past human use of landscapes is investigating how people adapted their behaviours to the different landscapes they inhabited. Often this kind of analyses depends on comparing modal attributes of archaeological sites between discrete patches of landscapes (e.g. Barton 2003). Variations in human behavior at landscape boundaries are more challenging to investigate, despite their importance in understanding sensitivity to landscape differences. Previous work on past human behaviors at ecological interfaces has focused on regions of highly contrasting biomes. For example, Epp (1985) found that archaeological sites were more concentrated in an ecotone (i.e. ecological transition area) where plains grasslands meet the boreal forest in Saskatchewan, compared to within those biomes. Epp’s work is typical of much early archaeological work on ecotones, with claims that ecotonal areas contain higher populations and diversities of flora and fauna (Struever 1968, Glassow and Leone 1972 ). This increased variety and density of plants and animals, often described as an ‘edge effect’, was claimed by Harris et al. (1969) to be a causal factor in the emergence of agriculture. This edge effect was also extended to human culture, with Gummerman and Johnson (1971) arguing for increased human cultural diversity in ecotonal zones of central Arizona.

@@

Our study area at Olympic Dam, South Australia, offers a unique opportunity to study landscape interfaces for two key reasons. First, in our study area these interfaces are relatively well-defined by landform attributes, such as the spacing between sand dune ridges and the surface geology (TABLE 1). The interfaces in our study area can be recognized as linear boundaries, often following the ridges or swales of the sand dunes, making them convenient for spatial analysis of archaeological features. Landform boundary effects have previously been employed in studying the distribution of artefacts to infer culture group areas (Kimes, Haselgrove et al. 1982). However, archaeological discussions of settlement patterns have typically focused on straight line distances from key resources such as raw materials for stone artefacts, as these are often easier to measure and interpret than landscape boundaries (Daniel 2001, Jones, Beck et al. 2003). With clear landscape unit boundaries, our study area is well-placed to explore relationships between archaeological attributes and distances from these boundaries.

@@

Hughes and Hiscock’s archaeological model has been described in detail in previous publications (Hughes et al. 2011; Hughes and Hiscock 1982; Hughes and Sullivan 1984). In brief, the model details how the location and character of archaeological sites are influenced by proximity to fresh water, sand on which to camp, and locally available raw materials for flaking stone artefacts. The model results in a series of predictions about the archaeological record expected for each of six landform types (TABLE 1) and five geological regimes (TABLE 2). These predictions specify the frequency of sites and the density and diversity of artefacts were produced for each major landscape unit.

These descriptions were initially tested in the 1980s and 1990s by archaeological surveys of the mining area and infrastructure corridors. The predictive model was based on records from an initial 133 sites and by 2007 additional testing and surveys of water and power supply corridors had resulted in a database totaling 820 archaeological sites  (Hiscock and Hughes 1983, Hiscock 1989).  In 2007 a proposed expansion of the Olympic Dam mine triggered the need for an intensive archaeological survey of an additional area of 515 km2 surrounding the existing mine, to be completed by the end of 2009. The data collected during this expansion phase is the basis for this study (Figure \@ref(fig:mapofstudyarea2)).

@@

Between 2007 and 2009 teams of archaeologists and trainees from local Aboriginal groups walking across the entire area recorded more than 16,000 archaeological sites. We achieved close to 90% coverage of 353 km2. The sites consist of stone artefact scatters (`r artefact_scatters`%), artefact scatters with knapping floors (`r as_kf`%), knapping floors (`r knapping_floors`%) and quarries (`r quarries`%) (Hughes, Hiscock et al. 2011). Extended descriptions of site definitions and survey methods have previously been presented in Hughes et al. (2011).

To collect field data, we used a mobile geographic information system (GIS) on handheld computers to record both background environmental and cadastral information and archaeological information. Our equipment was similar to what is commonly used during field survey (McPherron and Dibble 2003, Bevan and Conolly 2004, Tripcevich 2004, 2004, Given and Hyla 2006, Wagtendonk and De Jeu 2007, de la Vega and Agulla 2010, Tripcevich and Wernke 2010, Fei 2011, Scianna and Villa 2011, Traviglia 2011, Fee, Pettegrew et al. 2013, Newhard, Levine et al. 2013, Cascalheira, Gonçalves et al. 2014, Banning and Hitchings 2015). A detailed description of our specific hardware, software, and justifications for our choices is presented in our SOM.

We investigated behaviors at landform boundaries by analyzing the contrast between archaeological sites close to the boundary of a landform and the archaeology deeper within the landform. Our test of boundary effects focused on the archaeological evidence from resource-rich landform types/terrain patterns, such as Czs4, bordered by landscapes with lower resource levels, such as landform types 5 or 6. Because Czs4 and Czs5 dominate the 353 km2 sample area and contain the majority of sites, the analyses focused on comparing these two terrain patterns.

Two approaches were used to diagnose the presence and character of boundary effects. In the first approach, we determined the straight-line distance from every site to the nearest boundary of the terrain pattern in which it was contained. Descriptive statistics for these distances allows testing of the prediction that sites tend to be closer to the edge of some landforms than other landforms. In the second approach, we measured the density of sites at 100 m intervals, or buffer zones, inwards from the boundary of the terrain patterns Csz4 and Csz5, as the best represented terrain patterns. These data were used to test the prediction that site densities tends to be higher towards the center of Csz4, as a resource-rich landform, compared to Csz5, as a resource-poor landform. We created interior buffer polygons at 0-100 m, 101-200 m and 201-300 m from the boundary of the terrain patterns and the number of sites per square kilometer in each buffer was measured. The density of sites in each buffer zone was then divided by the average density for the entire terrain pattern to standardize the density value for the terrain pattern. Finally, z-scores were calculated to put the differences in site density on a comparable scale between the two terrain patterns. Z-scores are a transformation of the data that show how many standard deviations the site density in each buffer polygon varies from the mean density (z = 0) of the buffers in each terrain pattern. This second approach provided a way to minimize the effect of differences in the sizes of landscape units being compared. It does this by limiting analysis to an area 300 m from the boundary of the terrain pattern, thereby excluding sites in the middle of very large area terrain patterns.

# Results

@@

A wide variety of landforms and geological regimes were recorded, displayed in figure \@ref(fig:percentcoverageofeachterrainpattern) as the percentage coverage of each terrain pattern. In total these areas contained about `r all_sites_in_study_area` archaeological sites. Most of the study area consists of widely spaced dunes – landform type 4 (`r LF4`%), moderately spaced dunes – landform type 5 (`r LF5`%) and closely spaced dunes – landform type 6 (`r LF6`%), mainly overlying Tertiary stony-plains (Czs4 and Czs5 respectively) but some overlying limestone (A) or quartzite surfaces (K and P). Other terrain patterns make up less than 10% of the 353 km2 study area (Figure \@ref(fig:percentcoverageofeachterrainpattern)).

@@

About `r round(sites_in_Czs4 + sites_in_Czs5,0)`% of all sites were recorded in these two terrain patterns (Czs4: `r sites_in_Czs4`%, Czs5: `r sites_in_Czs5`%) with no other terrain pattern having more than `r sites_in_other_TPs`% of all sites. When considered as homogenous units, there are marked differences in the frequency of archaeological sites and the density of stone artefacts between these two terrain patterns. As predicted (TABLE 1), sites are more frequent in fields of widely spaced dunes (Czs4) with `r site_density_in_Czs4` sites/km2 compared with more closely spaced dunes (Czs5) which have only `r site_density_in_Czs5` sites/km2 (t(`r t_test_for_two_TPs_df_value`) = `r t_test_for_two_TPs_t_value`, p-value = `r t_test_for_two_TPs_p_value`) (Figure \@ref(fig:sitesinterrainpatterns). Also as predicted by the original model the density of surface-visible stone artefacts was substantially higher in areas of widely spaced dunes (Czs4), at `r artefacts_per_sqkm_Czs4` artefacts/km2, compared with closely spaced dunes Czs5 at `r artefacts_per_sqkm_Czs5` artefacts/km2.

@@

ODX_allsites_2010.s <- subset(ODX_allsites_2010, GREGLANDF=='A4' | GREGLANDF=='Czs4' | GREGLANDF=='Q4' | GREGLANDF=='K4' | GREGLANDF=='Czs5' | GREGLANDF=='Q5' | GREGLANDF=='Q6')
ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF)
# order the levels of GREGLANDF for plotting
# this is the desired order: - A4, Czs4, Q4, K4, Czs5, Q5, Q6
levels(ODX_allsites_2010.s$GREGLANDF)
ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF, 
  levels(ODX_allsites_2010.s$GREGLANDF)[c(1, 2, 5, 4, 3, 6, 7)])

@@

# now check the boundary densities for INNER buffers
inbd.files <- list.files("../data/inner_buffers", full.names = TRUE)
# list in, dataframe out
library(plyr)
library(foreign)
inbd.data <- ldply(inbd.files, read.dbf)
inbd.data$SiteDens <- with(inbd.data, Join_Count/buf_area * (1000*1000))
inbd.data <- inbd.data[inbd.data$Join_Count > n, ]
unique(inbd.data$BUFF_DIST)
inbd.data$Buf <- as.factor(with( inbd.data, ifelse(is.na(BUFF_DIST), '-100-000', ifelse(BUFF_DIST == -100, '-200-100', ifelse(BUFF_DIST == -200, '-300-200', ifelse(BUFF_DIST == -300, '-400-300', ifelse(BUFF_DIST == -400, '-500-400','NA')))))))

@@

Figure \@ref(fig:sitesinbuffers) shows that in Cz4 landscapes the density of sites increases toward the center of the landform while in Cz5 landforms site density decreases towards the center of the landform. The results of a two-way ANOVA show that the interaction of terrain patterns and the distances from the boundary have a significant effect on the density of archaeological sites at increasing distance from the boundary of the terrain pattern units (F(`r two_way_anova_df`) = `r two_way_anova_F`, p = `r two_way_anova_p`, excluding the 0-100 m zone). When we use z-scores to standardize for differences in site density between two terrain patterns, we see that the pattern is even more stark. Figure \@ref(fig:zscoresonsitedensitybyzones) shows that in Cz4 landscapes the density of sites increases toward the center of the landform while in Cz5 landforms site density decreases towards the center of the landform. This also minimizes the effects of differences in the sizes of landscape units by limiting analysis to an area 300 m from the boundary of the terrain pattern, thereby excluding sites in the middle of very large area terrain patterns. These analyses give support to the prediction that resource-rich landscapes such as Csz4 have sites located away from their margin and far inside the terrain pattern areas while landscapes with lower resource levels such as Csz5 commonly have sites closer to their margins and fewer sites further in their interiors.

@@

Our analysis of boundary effects shows people were senstive to subtle gradients of change in the landscape. We found that sites were generally located futher from the terrain pattern boundary in landform type 4, compared to landform types 5 and 6. This indicates that people preferred to be located close to the center of areas with widely-spared dunes, and on the edges of the areas with closely spaced dunes. Looking at this in more detail, we found that for Czs4 and Czs5 there is a clear contrast in the density of sites when we compare areas close to the middle of the terrain patterns with areas further away from the boundary of the terrain patterns.  
This analysis suggests that prehistoric people in the Olympic Dam area were sensitive to fine-grained differences in the economic potential of dissimilar terrain patterns. This claim is axiomatic to human behavioural ecology in which different landscapes are commonly compared as homogenous units to improve analytical tractability. The novel element of the data presented here is that the distribution of human activities varies substantially and predictably within as well as between terrain patterns. It is also noteworthy that a relatively subtle geographic differences between landscapes – the different spacing of sand dunes between Czs4 and Czs5 – has a marked visible effect on the distribution of archaeological sites. We speculate that this is due to the relationship between dune spacing and the availability of flakable stone. The causes of variation in dune spacing are poorly understood. Previous work has associated sand transport rates and directions, wind regime characteristics, grain size patterns, topography, and substrates with dune spacing, however there are no simple relationships (Lancaster 1988; Wasson and Hyde 1983; Wasson et al. 1988).

@@

Our survey activity in the Olympic Dam region over multiple decades has provided us with an opportunity to collect longitudinal data on site recording methods. Here we briefly explore the question of whether our mobile GIS enabled faster site recording then when we used a paper-pen-GPS method (these methods are described in more detail in our SOM). TABLE 5 shows a summary of the productivity of archaeological survey before and after we implemented our mobile GIS. Although there are many possible metrics of productivity, one that was especially relevant for our project was the number of sites recorded per person per day. This metric is useful because it is easy to measure without burdening field workers with additional data-logging tasks. However, we need to take into account possible confounding factors, such as travel time to the day’s survey area, the size and density of the sites in the day’s survey area, skill and efficiency of the individual archaeologist and the sampling strategy.

@@

@@

# linear model for sites recorded by density of sites
mobile_gis_efficiency_lm <- summary(lm(dat_t[,5] ~ dat_t[,6]))

# subset 2007-2009 data
dat_t_s <- dat_t[2:4,]
# linear model for sites recorded by density of sites, 2007-2009
mobile_gis_efficiency_lm_2007_2009 <- summary(lm(dat_t_s[,5] ~ dat_t_s[,6]))

# predict given 19 sites/km2
19 * 0.297  + 0.043 
```

Our pre-2007 data were collected by more experienced archaeologists than the 2007-2009 data, using a recording form with less variables than our mobile GIS recording forms, and the pre-2007 survey areas were all closer to the camp, with less travel time compared to the 2007-2009 data. Further, the average site size for the pre-2007 data is less than the 2007-2009 data. These conditions should favor a higher rate of recording for the 2007-2009 data. TABLE 5 shows that prior to adopting the mobile GIS in 2007 we recorded 3.5 sites per person per day. With the mobile GIS our rate increased substantially to an average of 10.2 sites per person per day for 2007-2009.

While this is a substantial difference, not all of this increase in the rate of recording can be attributed to the use of the mobile GIS because the rate of recording is strongly correlated with the density of sites (r = 0.97, p = 0.003). To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data (rate = site density * 0.297 (0.004)+ 0.062139 (0.146),  r2 = 0.999, p = 0.0002, standard errors in parentheses) and predicted the site recording rate given the density of 19 sites km-2 observed in the pre-2007 data. The resulting rate is 5.7 sites/person/day, which is what we would expect if a mobile GIS had been used to record sites in an area with 19 sites km-2. The difference of 2.2 sites/person/day between the actual pre-2007 average rate of 3.5 and the extrapolated average rate of 5.7 is most likely due to the use of mobile GIS. This represents a sixty percent increase in the site recording rate. This demonstrates an advantage for using mobile GIS in a context where data collection was intensive, a highly structured data collection form could be used, and the start-up and training costs could be amortized over multiple years.

@@

This work was undertaken under a contract with BHP Billiton that includes research support. Other financial support for Ben Marwick came from a Luce/ACLS Post-doctoral fellowship (B9471 01 180401 7301) and a grant from the Herbert H. Gowen International Studies Endowment (UW Office of Global Affairs).

>>>>

Philip Hughes    
Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia

Marjorie Sullivan    
Huonbrook and Environment and Heritage P/L, PO Box 97 Moruya NSW 2537, Australia

@@

A key step in understanding past human use of landscapes is investigating how people adapted their behaviors to the different landscapes they inhabited. Often this kind of analyses depends on comparing modal attributes of archaeological sites between discrete patches of landscapes (e.g. Barton 2003). Variations in human behavior at landscape boundaries are more challenging to investigate, despite their importance in understanding sensitivity to landscape differences. Previous work on past human behaviors at ecological interfaces has focused on regions of highly contrasting biomes. For example, Epp (1985) found that archaeological sites were more concentrated in an ecotone (i.e. ecological transition area) where plains grasslands meet the boreal forest in Saskatchewan, compared to within those biomes. Epp’s work is typical of much early archaeological work on ecotones, with claims that ecotonal areas contain higher populations and diversities of flora and fauna (Struever 1968, Glassow and Leone 1972). This increased variety and density of plants and animals, often described as an ‘edge effect’, was claimed by Harris et al. (1969) to be a causal factor in the emergence of agriculture. This edge effect was also extended to human culture, with Gummerman and Johnson (1971) arguing for increased human cultural diversity in ecotonal zones of central Arizona.

@@

##### . 
<!-- page break -->

```{r table1}
table1 <- read.csv("../data/table_1.csv")
names(table1) <- c("Landform type", "Archaeological prediction")
kable(table1, caption = "Landform types and predicted archaeological signature. Adapted from Hughes et al. (2011), Kinhill-Stearns Roger (1982) and Hughes and Sullivan (1984).")
```

Our study area at Olympic Dam, South Australia, offers a unique opportunity to study landscape interfaces for two key reasons. First, in our study area these interfaces are relatively well-defined by landform attributes, such as the spacing between sand dune ridges and the surface geology (Table \@ref(tab:table1)). The interfaces in our study area can be recognized as linear boundaries, often following the ridges or swales of the sand dunes, making them convenient for spatial analysis of archaeological features. Landform boundary effects have previously been employed in studying the distribution of artefacts to infer culture group areas (Kimes, Haselgrove et al. 1982). However, archaeological discussions of settlement patterns have typically focused on straight line distances from key resources such as raw materials for stone artefacts, as these are often easier to measure and interpret than landscape boundaries (Daniel 2001, Jones, Beck et al. 2003). With clear landscape unit boundaries, our study area is well-placed to explore relationships between archaeological attributes and distances from these boundaries.

@@

##### . 
<!-- page break -->

```{r table2}
table2 <- read.csv("../data/table_2.csv")
names(table2) <- c("Geological regime", "Period", "Description", "Materials for artefact manufacture")
kable(table2, caption = "Geological regimes in the study area and predicted archaeological importance. Adapted from Hughes et al. (2011).")
```

Hughes and Hiscock’s archaeological model has been described in detail in previous publications (Hughes et al. 2011; Hughes and Hiscock 1982; Hughes and Sullivan 1984). In brief, the model details how the location and character of archaeological sites are influenced by proximity to fresh water, sand on which to camp, and locally available raw materials for flaking stone artefacts. The model results in a series of predictions about the archaeological record expected for each of six landform types (Table \@ref(tab:table1)) and five geological regimes (Table \@ref(tab:table2)). These predictions specify the frequency of sites and the density and diversity of artefacts were produced for each major landscape unit.

These descriptions were initially tested in the 1980s and 1990s by archaeological surveys of the mining area and infrastructure corridors. The predictive model was based on records from an initial 133 sites and by 2007 additional testing and surveys of water and power supply corridors had resulted in a database totaling 820 archaeological sites  (Hiscock and Hughes 1983, Hiscock 1989).  In 2007 a proposed expansion of the Olympic Dam mine triggered the need for an intensive archaeological survey of an additional area of 515 km^2^ surrounding the existing mine, to be completed by the end of 2009. The data collected during this expansion phase is the basis for this study (Figure \@ref(fig:mapofstudyarea2)).

@@

Between 2007 and 2009 teams of archaeologists and trainees from local Aboriginal groups walking across the entire area recorded more than 16,000 archaeological sites. We achieved close to 90% coverage of 353 km^2^. The sites consist of stone artefact scatters (`r artefact_scatters`%), artefact scatters with knapping floors (`r as_kf`%), knapping floors (`r knapping_floors`%) and quarries (`r quarries`%) (Hughes, Hiscock et al. 2011). Extended descriptions of site definitions and survey methods have previously been presented in Hughes et al. (2011).

To collect field data, we used a mobile geographic information system (GIS) on handheld computers to record both background environmental and cadastral information and archaeological information. Our equipment was similar to what is commonly used during field survey (McPherron and Dibble 2003, Bevan and Conolly 2004, Tripcevich 2004, 2004, Given and Hyla 2006, Wagtendonk and De Jeu 2007, de la Vega and Agulla 2010, Tripcevich and Wernke 2010, Fei 2011, Scianna and Villa 2011, Traviglia 2011, Fee, Pettegrew et al. 2013, Newhard, Levine et al. 2013, Cascalheira, Gonçalves et al. 2014, Banning and Hitchings 2015). Our mobile GIS configuration files, recording forms, and scripts are archived online at <http://doi.org/10.5281/zenodo.11833>. A detailed description of our specific hardware, software, and justifications for our choices is presented in our SOM.

### Investigating boundary effects

We investigated behaviors at landform boundaries by analyzing the contrast between archaeological sites close to the boundary of a landform and the archaeology deeper within the landform. Our test of boundary effects focused on the archaeological evidence from resource-rich landform types/terrain patterns, such as Czs4, bordered by landscapes with lower resource levels, such as landform types 5 or 6. Because Czs4 and Czs5 dominate the 353 km^2^ sample area and contain the majority of sites, the analyses focused on comparing these two terrain patterns.

Two approaches were used to diagnose the presence and character of boundary effects. In the first approach, we determined the straight-line distance from every site to the nearest boundary of the terrain pattern in which it was contained. Descriptive statistics for these distances allows testing of the prediction that sites tend to be closer to the edge of some landforms than other landforms. In the second approach, we measured the density of sites at 100 m intervals, or buffer zones, inwards from the boundary of the terrain patterns Csz4 and Csz5, as the best represented terrain patterns. These data were used to test the prediction that site densities tends to be higher towards the center of Csz4, as a resource-rich landform, compared to Csz5, as a resource-poor landform. We created interior buffer polygons at 0-100 m, 101-200 m and 201-300 m from the boundary of the terrain patterns and the number of sites per square kilometer in each buffer was measured. The density of sites in each buffer zone was then divided by the average density for the entire terrain pattern to standardize the density value for the terrain pattern.

Finally, z-scores were calculated to put the differences in site density on a comparable scale between the two terrain patterns. Z-scores are a transformation of the data that show how many standard deviations the site density in each buffer polygon varies from the mean density (z = 0) of the buffers in each terrain pattern. This second approach provided a way to minimize the effect of differences in the sizes of landscape units being compared. It does this by limiting analysis to an area 300 m from the boundary of the terrain pattern, thereby excluding sites in the middle of very large area terrain patterns.

### Reproducibility and open source materials

To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in Marwick (2016), we include the entire R code used for all the analysis and visualizations contained in this paper in our SOM at <http://dx.doi.org/xxx/xxx>. Also in this version-controlled compendium are the raw data for all the tests reported here, as well as additional regression diagnostics and power tests. All of the figures, tables and statistical test results presented here can be independently reproduced with the code and data in this repository. In our SOM our code is released under the MIT licence, our data as CC-0, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2016).

## Results

@@

A wide variety of landforms and geological regimes were recorded, displayed in figure \@ref(fig:percentcoverageofeachterrainpattern) as the percentage coverage of each terrain pattern. In total these areas contained about `r all_sites_in_study_area` archaeological sites. Most of the study area consists of widely spaced dunes – landform type 4 (`r LF4`%), moderately spaced dunes – landform type 5 (`r LF5`%) and closely spaced dunes – landform type 6 (`r LF6`%), mainly overlying Tertiary stony-plains (Czs4 and Czs5 respectively) but some overlying limestone (A) or quartzite surfaces (K and P). Other terrain patterns make up less than 10% of the 353 km^2^ study area (Figure \@ref(fig:percentcoverageofeachterrainpattern)).

@@

About `r round(sites_in_Czs4 + sites_in_Czs5,0)`% of all sites were recorded in these two terrain patterns (Czs4: `r sites_in_Czs4`%, Czs5: `r sites_in_Czs5`%) with no other terrain pattern having more than `r sites_in_other_TPs`% of all sites. When considered as homogeneous units, there are marked differences in the frequency of archaeological sites and the density of stone artefacts between these two terrain patterns. As predicted (Table \@ref(tab:table1)), sites are more frequent in fields of widely spaced dunes (Czs4) with `r site_density_in_Czs4` sites/km^2^ compared with more closely spaced dunes (Czs5) which have only `r site_density_in_Czs5` sites/km^2^ (t(`r t_test_for_two_TPs_df_value`) = `r t_test_for_two_TPs_t_value`, p-value = `r t_test_for_two_TPs_p_value`) (Figure \@ref(fig:sitesinterrainpatterns). Also as predicted by the original model the density of surface-visible stone artefacts was substantially higher in areas of widely spaced dunes (Czs4), at `r artefacts_per_sqkm_Czs4` artefacts/km^2^, compared with closely spaced dunes Czs5 at `r artefacts_per_sqkm_Czs5` artefacts/km^2^.

@@

ODX_allsites_2010.s <- subset(ODX_allsites_2010, GREGLANDF=='A4' | GREGLANDF=='Czs4' | GREGLANDF=='Q4' | GREGLANDF=='K4' | GREGLANDF=='Czs5' | GREGLANDF=='Q5' | GREGLANDF=='Q6')
ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF)
# order the levels of GREGLANDF for plotting
# this is the desired order: - A4, Czs4, Q4, K4, Czs5, Q5, Q6
# levels(ODX_allsites_2010.s$GREGLANDF)
ODX_allsites_2010.s$GREGLANDF <- factor(ODX_allsites_2010.s$GREGLANDF, 
  levels(ODX_allsites_2010.s$GREGLANDF)[c(1, 2, 5, 4, 3, 6, 7)])

@@

# now check the boundary densities for INNER buffers
inbd.files <- list.files("../data/inner_buffers", full.names = TRUE)
# list in, dataframe out
library(plyr)
library(foreign)
inbd.data <- ldply(inbd.files, read.dbf)
inbd.data$SiteDens <- with(inbd.data, Join_Count/buf_area * (1000*1000))
inbd.data <- inbd.data[inbd.data$Join_Count > n, ]
# unique(inbd.data$BUFF_DIST)
inbd.data$Buf <- as.factor(with( inbd.data, ifelse(is.na(BUFF_DIST), '-100-000', ifelse(BUFF_DIST == -100, '-200-100', ifelse(BUFF_DIST == -200, '-300-200', ifelse(BUFF_DIST == -300, '-400-300', ifelse(BUFF_DIST == -400, '-500-400','NA')))))))

@@

Figure \@ref(fig:sitesinbuffers) shows that in Czs4 landscapes the density of sites increases toward the center of the landform while in Czs5 landforms site density decreases towards the center of the landform. The results of a two-way ANOVA show that the interaction of terrain patterns and the distances from the boundary have a significant effect on the density of archaeological sites at increasing distance from the boundary of the terrain pattern units (F(`r two_way_anova_df`) = `r two_way_anova_F`, p = `r two_way_anova_p`, excluding the 0-100 m zone). When we use z-scores to standardize for differences in site density between two terrain patterns, we see that the pattern is even more stark. Figure \@ref(fig:zscoresonsitedensitybyzones) shows that in Czs4 landscapes the density of sites increases toward the center of the landform while in Czs5 landforms site density decreases towards the center of the landform. This also minimizes the effects of differences in the sizes of landscape units by limiting analysis to an area 300 m from the boundary of the terrain pattern, thereby excluding sites in the middle of very large area terrain patterns. These analyses give support to the prediction that resource-rich landscapes such as Csz4 have sites located away from their margin and far inside the terrain pattern areas while landscapes with lower resource levels such as Csz5 commonly have sites closer to their margins and fewer sites further in their interiors.

@@

Our analysis of boundary effects shows people were sensitive to subtle gradients of change in the landscape. We found that sites were generally located further from the terrain pattern boundary in landform type 4, compared to landform types 5 and 6. This indicates that people preferred to be located close to the center of areas with widely-spared dunes, and on the edges of the areas with closely spaced dunes. Looking at this in more detail, we found that for Czs4 and Czs5 there is a clear contrast in the density of sites when we compare areas close to the middle of the terrain patterns with areas further away from the boundary of the terrain patterns.

This analysis suggests that prehistoric people in the Olympic Dam area were sensitive to fine-grained differences in the economic potential of dissimilar terrain patterns. This claim is axiomatic to human behavioral ecology in which different landscapes are commonly compared as homogeneous units to improve analytical tractability. The novel element of the data presented here is that the distribution of human activities varies substantially and predictably within as well as between terrain patterns. It is also noteworthy that a relatively subtle geographic differences between landscapes – the different spacing of sand dunes between Czs4 and Czs5 – has a marked visible effect on the distribution of archaeological sites. We speculate that this is due to the relationship between dune spacing and the availability of flakable stone. The causes of variation in dune spacing are poorly understood. Previous work has associated sand transport rates and directions, wind regime characteristics, grain size patterns, topography, and substrates with dune spacing, however there are no simple relationships (Lancaster 1988; Wasson and Hyde 1983; Wasson et al. 1988).

@@

Our survey activity in the Olympic Dam region over multiple decades has provided us with an opportunity to collect longitudinal data on site recording methods. Here we briefly explore the question of whether our mobile GIS enabled faster site recording then when we used a paper-pen-GPS method (these methods are described in more detail in our SOM). Table \@ref(tab:mobilegisefficiency) shows a summary of the productivity of archaeological survey before and after we implemented our mobile GIS. Although there are many possible metrics of productivity, one that was especially relevant for our project was the number of sites recorded per person per day. This metric is useful because it is easy to measure without burdening field workers with additional data-logging tasks. However, we need to take into account possible confounding factors, such as travel time to the day’s survey area, the size and density of the sites in the day’s survey area, skill and efficiency of the individual archaeologist and the sampling strategy.

@@

display_table <- dat
display_table$"" <- row.names(display_table)
row.names(display_table) <- NULL
display_table <- display_table[ , c(5, 1:4)]
names(display_table) <- c("", "pre-2007", "2007", "2008", "2009")
display_table[ , 1] <- gsub("\\.", " ", display_table[ , 1])
display_table[ , 1] <- gsub("km2", "km^2^", display_table[ , 1])
display_table[ , 1] <- gsub("persondays", "person-days", display_table[ , 1])

# print table
kable(display_table, caption = "Summary of archaeological survey productivity at Olympic Dam. The data for sites recorded per person per day for before 2007 come from a survey undertaken in 1982 by Philip Hughes and Peter Hiscock that recorded 176 sites in 50 person days.")

# modelling these data

@@

# rate for pre-2007 and 2007-2009
rate_for_pre_2007 <- dat_t$No.sites.recorded.person.day[1]  
rate_for_2007_thru_2009 <-  round(mean(dat_t$No.sites.recorded.person.day[c(2:4)]),1)

# the rate of recording is strongly correlated with the density of sites 
# linear model for sites recorded by density of sites
mobile_gis_efficiency_lm <- 
  summary(mobile_gis_efficiency_lm_out <- lm(dat_t$No.sites.recorded.person.day ~ dat_t$Frequency.of.occurrence.of.sites.km2))
mobile_gis_efficiency_lm_r_squared <- round(mobile_gis_efficiency_lm$adj.r.squared, 3)
mobile_gis_efficiency_lm_p_value <- round(mobile_gis_efficiency_lm$coefficients[2, 4]  , 3)

# a linear model for the 2007-2009 data
# subset 2007-2009 data
dat_t_s <- dat_t[2:4,]
# linear model for sites recorded by density of sites, 2007-2009
mobile_gis_efficiency_lm_2007_2009 <-
 summary(mobile_gis_efficiency_lm_2007_2009_out <- 
 lm(dat_t_s$No.sites.recorded.person.day ~ dat_t_s$Frequency.of.occurrence.of.sites.km2))

# display model
mobile_gis_efficiency_lm_2007_2009_est <- round(mobile_gis_efficiency_lm_2007_2009$coefficients[2,1], 3)
mobile_gis_efficiency_lm_2007_2009_est_error <-  round(mobile_gis_efficiency_lm_2007_2009$coefficients[2,2], 3)
mobile_gis_efficiency_lm_2007_2009_inter <-  round(mobile_gis_efficiency_lm_2007_2009$coefficients[1,1], 3)
mobile_gis_efficiency_lm_2007_2009_inter_error <-   round(mobile_gis_efficiency_lm_2007_2009$coefficients[1,2], 3)
mobile_gis_efficiency_lm_2007_2009_r_sq_adj <- round(mobile_gis_efficiency_lm_2007_2009$adj.r.squared, 3)
mobile_gis_efficiency_lm_2007_2009_p_value <- round(mobile_gis_efficiency_lm_2007_2009$coefficients[2, 4], 3)

# predict given 19 sites/km2 of pre-2007
resulting_rate <- round(19 * mobile_gis_efficiency_lm_2007_2009_est + mobile_gis_efficiency_lm_2007_2009_inter, 2)

# difference in rates, pre-2007 and 2007-2009
difference_in_rates_pre_2007_and_2007_2009 <- 
resulting_rate - rate_for_pre_2007

# percentage increase
percentage_increase <- round((resulting_rate - rate_for_pre_2007) / rate_for_pre_2007 * 100, 0)
```

Our pre-2007 data were collected by more experienced archaeologists than the 2007-2009 data, using a recording form with less variables than our mobile GIS recording forms, and the pre-2007 survey areas were all closer to the camp, with less travel time compared to the 2007-2009 data. Further, the average site size for the pre-2007 data is less than the 2007-2009 data. These conditions should favor a higher rate of recording for the 2007-2009 data. Table \@ref(tab:mobilegisefficiency) shows that prior to adopting the mobile GIS in 2007 we recorded `r rate_for_pre_2007` sites per person per day. With the mobile GIS our rate increased substantially to an average of `r rate_for_2007_thru_2009` sites per person per day for 2007-2009.

While this is a substantial difference, not all of this increase in the rate of recording can be attributed to the use of the mobile GIS because the rate of recording is strongly predicted by the density of sites (R^2^~adj~ = `r mobile_gis_efficiency_lm_r_squared`, p = `r mobile_gis_efficiency_lm_p_value`). To isolate the effect of site density on recording rate we computed a linear model for the 2007-2009 data , which resulted in a rate of site recording that can by predicted by site density * `r mobile_gis_efficiency_lm_2007_2009_est` (`r mobile_gis_efficiency_lm_2007_2009_est_error`) + `r mobile_gis_efficiency_lm_2007_2009_inter` (`r mobile_gis_efficiency_lm_2007_2009_inter_error`) (standard errors in parentheses). This model gave a good fit (R^2^~adj~ = `r mobile_gis_efficiency_lm_2007_2009_r_sq_adj`, p = `r mobile_gis_efficiency_lm_2007_2009_p_value`), and was then used to predict the site recording rate given the density of 19 sites/km^2^ observed in the pre-2007 data. The resulting rate is `r resulting_rate` sites/person/day, which is what the model predicts if a mobile GIS had been used to record sites in an area with 19 sites/km^2^. The difference of `r difference_in_rates_pre_2007_and_2007_2009` sites/person/day between the actual pre-2007 average rate of `r rate_for_pre_2007` and the predicted average rate of `r resulting_rate` is most likely due to the use of mobile GIS. This represents a `r percentage_increase`% increase in the site recording rate, although we caution that this may be an overestimate due to overfitting from the relatively small number of data points in our efficiency analysis. Nevertheless, this demonstrates an advantage for using mobile GIS in a context where data collection was intensive, a highly structured data collection form could be used, and the start-up and training costs could be amortized over multiple years.

@@

This work was undertaken under a contract with BHP Billiton that includes research support. Other financial support for Ben Marwick came from a Luce/ACLS Post-doctoral fellowship (B9471 01 180401 7301), a grant from the Herbert H. Gowen International Studies Endowment (UW Office of Global Affairs), and an Australian Research Council Future Fellowship (FT140100101).
